---
title: "Estimating the Causal Effect of Pandemic-Induced League Shutdowns on Hockey Prospect Development"
subtitle: "Background Walk-Through"
author: "Jackie Jovanovic"
output: 
  html_document:
    code_folding: hide
date: 'September 13th, 2022'
---

<link rel="preconnect" href="https://fonts.googleapis.com">
<link rel="preconnect" href="https://fonts.gstatic.com" crossorigin>
<link href="https://fonts.googleapis.com/css2?family=Open+Sans:wght@300&family=Source+Sans+Pro&display=swap" rel="stylesheet">
<style>

body{
font-family: 'Source Sans Pro', sans-serif;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, include=FALSE}
# load packages
library(tidyverse)
library(knitr)
```

## Load the original data and explore:
```{r load_data, include=FALSE}
library(readr)
sams_ohl_data_request <- read_csv("sams_ohl_data_request.csv")
```

Variables:
```{r variables}
names(sams_ohl_data_request)
```

```{r, message = FALSE, warning = FALSE, echo=FALSE}
sams_ohl_data_request %>% 
  dplyr::select(name, team_name, season, league, position, 
                gp, g, a, pts, pm) %>%
  head(5) %>% 
  knitr::kable() %>% 
  kableExtra::kable_styling("striped")
```

```{r, include=FALSE}
# find number of seasons
length(unique(sams_ohl_data_request$season))

# find number of leagues
length(unique(sams_ohl_data_request$league))

# find number of players
length(unique(sams_ohl_data_request$name))

# drafted players
drafted_orig <- sams_ohl_data_request %>%
  filter(!is.na(draft_year)) 
length(unique(drafted_orig$name))
```

* 11690 observations
* 28 variables
* 40 seasons (earliest is 1982-1983, latest is 2021-2022)
* 159 leagues
* 3311 players
* 873 players drafted

## Data filtered to the three seasons of interest (2019-2020, 2020-2021, 2021-2022):

```{r, include=FALSE}
# filter years
recent <- sams_ohl_data_request %>% 
  filter(season %in% c("2019-2020", "2020-2021", "2021-2022"))

# find number of leagues
length(unique(recent$league))

# find number of players
length(unique(recent$name))

# drafted players
drafted_rec <- recent %>%
  filter(!is.na(draft_year)) 
length(unique(drafted_rec$name))
```

```{r, message = FALSE, warning = FALSE, echo=FALSE}
recent %>% 
  dplyr::select(name, team_name, season, league, position, 
                gp, g, a, pts, pm) %>%
  head(5) %>% 
  knitr::kable() %>% 
  kableExtra::kable_styling("striped")
```

* 2103 observations
* 28 variables
* 3 seasons
* 159 leagues
* 733 players
* 136 players drafted

```{r}
# players that appear in all three seasons
inall3 <- recent %>% 
  group_by(name) %>% 
  summarize(count = n())

# full stats for players that played in all three seasons
## filter(gene_ID %in% accessions40$V1) ## online example ##
inall3_full <- recent %>% 
  filter(name %in% inall3$name)

```

## How many seasons does one typically play in the OHL?
Using the entire original dataset (40 seasons)
```{r}
sams_ohl_data_request %>% 
  group_by(name) %>% 
  summarize(count = n())

# filter to only the OHL
only_ohl <- sams_ohl_data_request %>% 
  filter(league == "OHL")
  #unique(sams_ohl_data_request$name)

# find they amount of seasons they played in the OHL (count)
only_ohl_counts <- only_ohl %>% 
  group_by(name) %>% 
  summarize(count = n())

# summarize, find average 
ohl_counts_summary <- data.frame(Summary = c("Minimum:", "1st Quarter:",
                                             "Median:", "Mean:",
                                             "3rd Quarter:", "Maximum:"), 
                                 Value = c(1.000, 2.000, 3.000, 3.211,
                                           4.000, 12.000))
ohl_counts_summary %>% 
  kable() %>%
  kableExtra::kable_styling("striped")
```


