---
title: "Estimating the Causal Effect of Pandemic-Induced League Shutdowns on Hockey Prospect Development"
subtitle: "Background Walk-Through"
author: "Jackie Jovanovic"
output: html_document
date: 'September 20th, 2022'
---

<link rel="preconnect" href="https://fonts.googleapis.com">
<link rel="preconnect" href="https://fonts.gstatic.com" crossorigin>
<link href="https://fonts.googleapis.com/css2?family=Open+Sans:wght@300&family=Source+Sans+Pro&display=swap" rel="stylesheet">
<style>

body{
font-family: 'Source Sans Pro', sans-serif;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, include=FALSE}
# load packages
library(tidyverse)
library(readr)
library(knitr)
library(patchwork)

library(BART)
library(tidytreatment)
library(tidybayes)
```

## Load the original data and explore:
```{r, include=FALSE}
sams_ohl_data_request <- read_csv("sams_ohl_data_request.csv")
```

## Data filtered to the three seasons of interest (2019-2020, 2020-2021, 2021-2022):
```{r}
# filter years
recent <- sams_ohl_data_request %>% 
  filter(season %in% c("2019-2020", "2020-2021", "2021-2022"))
```

## ohl_filtered
```{r}
ohl_filtered <- read_csv("ohl_filtered.csv")

ohl_filtered <- ohl_filtered %>% 
  mutate(got_drafted = case_when(!is.na(draft_year) & draft_year < 2020
                                 ~ 'Yes',
                                 TRUE ~ 'No'))

# convert treatment to binary indicator:
ohl_filtered <- ohl_filtered %>% 
  mutate(treatment = ifelse(treatment == "Played", 1, 0),
         treatment = as.integer((treatment)))

# convert categorical variables to binary
ohl_update <- ohl_filtered %>% 
  mutate(is_forward = case_when(position == "F" ~ 1, 
                                position == "D" ~ 0)) %>% 
  mutate(is_drafted = case_when(got_drafted == "Yes" ~ 1,
                                got_drafted == "No" ~ 0)) %>% 
  mutate(s_played = case_when(season == "2021-2022" ~ 1,
                              season == "2019-2020" ~ 0)) %>% 
  as.data.frame()

```

Following the guidelines of <https://cran.r-project.org/web/packages/tidytreatment/vignettes/use-tidytreatment-BART.html>

### Fitting variable selection model

Regress outcome against covariates (excluding treatment variable)

Uses the **wbart** function: <https://www.rdocumentation.org/packages/BART/versions/2.9/topics/wbart>

variables included: games played, points, age, position, draft status

possible variables to include: championship (note to self: would need to convert to binary)
which pm variables should be included? : pm, pm_rank, pm_relative

```{r fit-var-sel, message = FALSE, warning = FALSE}
# fit variable selection model
var_select_bart <- wbart(x.train = dplyr::select(ohl_update, gp_19_20,
                                                 pts_19_20,age_continuous, 
                                                 pm_relative_19_20, pm_19_20, 
                                                 is_forward, is_drafted),
                       y.train = pull(ohl_update, ppg_19_20),
                       sparse = TRUE,
                       ntree = 25,
                       ndpost = 50)
```
I used ntree = 25 and ndpost = 50 because I was trying to follow the steps in <https://bayesiancomputationbook.com/markdown/chp_07.html#fitting-bayesian-additive-regression-trees>
And I figured the smaller size would make it less likely for R to crash.

review output
something with tree draws?

### Selecting subset of covariate from the VS model

select most important vars from y ~ covariates model

```{r var-sel, message = FALSE, warning = FALSE}
# variable selection
covar_ranking <- covariate_importance(var_select_bart)
var_select <- covar_ranking %>% 
  filter(avg_inclusion >= quantile(avg_inclusion, 0.5)) %>% 
  pull(variable)
```

covariate_importance description: "inclusion metric for bartMachine and BART are scaled different. bartMachine averaged over number of trees, in addition to number of MCMC draws."

### Fitting propensity score model

A probit/logit model estimating the propensity score using only the covariates selected in the previous step.
Regress z ~ selected covariates

Uses the **pbart** function: <https://www.rdocumentation.org/packages/BART/versions/2.9/topics/pbart>

```{r prop-scr-mdl, message = FALSE, warning = FALSE}
# fit a propensity score model
prop_bart <- pbart(x.train = dplyr::select(ohl_update, all_of(var_select)),
                  y.train = pull(ohl_update, treatment),
                  nskip = 2000,
                  ndpost = 5000)

# store propensity score in data
ohl_update$prop_score <- prop_bart$prob.train.mean
```

### Fitting the treatment effect model

Using the original covariates and propensity score from step 3.
Regress y ~ z + covariates + propensity score

```{r fit-trt-ef, message = FALSE, warning = FALSE}
# fit the treatment effect model
te_model <- wbart(x.train = dplyr::select(ohl_update, gp_19_20, pts_19_20,
                                  age_continuous, pm_relative_19_20,
                                  pm_rank_19_20, pm_19_20, is_forward,
                                  is_drafted, treatment, prop_score),
                                  # need to include treatment and prop score
                 y.train = pull(ohl_update, ppg_19_20),
                 nskip = 10000L,
                 ndpost = 200L,
                 keepevery = 100L)
```

Many of these functions produce data types that are exclusive to that specific function - the 'wbart' function returns the an object of type 'wbart', which is basically a list. Because of this I'm not sure what other functions (that are not in the tidytreatment outline) are compatible with these objects.

#### Question:
What data should I use for this? If I used the full original data, there will be the issue of multiple columns for one player in one season (if they played with multiple teams/leagues). The way I know how to fix that would cut out a lot of the variables. To combine player rows would I have to do summarize, for things like games played and points, then add the other columns (like name and draft year) back in. One problem with this method is I won't be able to look at league or team to draw conclusions concerning those things.

### Extract the posterior
```{r extratct-post, message = FALSE, warning = FALSE}
# extract the posterior
posterior_fitted <- fitted_draws(te_model, value = "fit",
                                 include_newdata = FALSE)
posterior_fitted

# function to tidy predicted draws and add random normal noise by default
posterior_pred <- predicted_draws(te_model, include_newdata = FALSE)
```
**fitted_draws** has been deprecated. It is from the tidybayes package. Here is the documentation: <https://www.rdocumentation.org/packages/tidybayes/versions/1.1.0/topics/add_fitted_draws>
The help menu gave other functions that can be used instead of fitted_draws, but I am not sure which one to use because I don't know what to values to use. I was going to try both functions out and compare the results to what was produced from fitted_draws but then I ran into the issue with the arguments.

```{r}
#add_epred_draws(posterior_pred, te_model)
# returns the error: "no applicable method for posterior_epred' applied ot an object of class 'wbart'"
```

### Plot with **tidybayes** {.tabset}

#### Position
```{r plot-tidyb-pos, message = FALSE, warning = FALSE, fig.align = 'center'}
# plots using tidybayes
treatment_var_and_c1 <- ohl_update %>% 
  dplyr::select(treatment, is_forward) %>% 
  mutate(.row = 1:n(), treatment = as.factor(treatment))

posterior_fitted %>% 
  left_join(treatment_var_and_c1, by = ".row") %>% 
  ggplot() +
  stat_halfeye(aes(x = treatment, y = fit)) +
  facet_wrap(~is_forward, labeller = 
               as_labeller( function(x) paste("is_forward =",x) ) ) +
  xlab("Treatment (played during covid)") + ylab("Posterior predicted value") +
  theme_bw() + 
  ggtitle("Effect of treatment with 'is_forward' on posterior fitted value")
```

#### Age
```{r plot-tidyb-age, message = FALSE, warning = FALSE, fig.align = 'center'}
# plots using tidybayes
treatment_var_and_c1 <- ohl_update %>% 
  dplyr::select(treatment, age) %>% 
  mutate(.row = 1:n(), treatment = as.factor(treatment))

posterior_fitted %>% 
  left_join(treatment_var_and_c1, by = ".row") %>% 
  ggplot() +
  stat_halfeye(aes(x = treatment, y = fit)) +
  facet_wrap(~age, labeller = 
               as_labeller( function(x) paste("age =",x) ) ) +
  xlab("Treatment (played during covid)") + ylab("Posterior predicted value") +
  theme_bw() + 
  ggtitle("Effect of treatment with 'age' on posterior fitted value")
```
I have no idea what's going on with age = 17

### {-}

#### Issue
Some players that have drafted = FALSE have a value in the draft_year column. It's only for about 20 players, the rest with drafted = FALSE have an 'NA' in the draft_year column. 

### Treatment effects

Posterior conditional (average) treatment effects calculated using **treatment_effects** function. There is an equation showing the equation in the "Calculate Treatment Effects" section of the tidytreatment page, I just don't know how to write equations in R.
```{r calc-trt-ef, message = FALSE, warning = FALSE}
# calculate treatment effects
# sample based (using data from fit) conditional treatment effects,
# posterior draws
posterior_treat_eff <- 
  treatment_effects(te_model, treatment = "treatment",
                    # the dataset here needs to match the BART data EXACTLY
                    # which is really annoying...
                    newdata = dplyr::select(ohl_update, gp_19_20, pts_19_20, 
                                     age_continuous, pm_relative_19_20, 
                                     pm_rank_19_20, pm_19_20, is_forward, 
                                     is_drafted, treatment, prop_score))
```
All draws:
```{r, trt_ef_all, message = FALSE, warning = FALSE, fig.align = 'center'}
# histogram of treatment effect (all draws)
posterior_treat_eff %>% 
  ggplot() +
  geom_histogram(aes(x = cte), bins = 50, color = "white") +
  theme_bw() + ggtitle("Histogram of treatment effect (all draws)")
```
This appears to be normally distributed as the bell-curve looks to be pretty symmetric. The right side may be a bit longer, but it is minimal.

Mean for each subject:
```{r trt_ef_sub, message = FALSE, warning = FALSE, fig.align = 'center'}
# histogram of treatment effect (mean for each subject)
posterior_treat_eff %>% summarise(cte_hat = mean(cte)) %>%
  ggplot() +
  geom_histogram(aes(x = cte_hat), bins = 60, colour = "white") + 
  theme_bw() + 
  ggtitle("Histogram of treatment effect (mean for each subject)")
```
This histogram is clearly very different from the first one. If only looking at the center chunk (ignoring the outliers for now), the plot seems fairly symmetric. There are a lot of outliers. One suggestion Sam had was looking through the data to find the outliers and figure out what was making them stick out. There are a couple issues with that. First, the dataset used for this graph is 43800 observations. Second, since the data was generated by the function there are no names associated with each observation. The only thing I could possibly use to backtrack it is the '.row' column, but the rows range from 1 to 219 so this approach is also unrealistic.

Confidence intervals of the CATEs:
stack of each subjects posterior confidence intervals of the CATEs
```{r ci-cate, message = FALSE, warning = FALSE, fig.align = 'center'}
# posterior CIs of the CATEs
posterior_treat_eff %>% select(-treatment) %>% point_interval() %>%
  arrange(cte) %>% mutate(.orow = 1:n()) %>% 
  ggplot() + 
  geom_interval(aes(x = .orow, y= cte, ymin = .lower, ymax = .upper)) +
  geom_point(aes(x = .orow, y = cte), shape = "circle open", alpha = 0.5) + 
  ylab("Median posterior CATE for each subject (95% CI)") +
  theme_bw() + coord_flip() + scale_colour_brewer() +
  theme(axis.title.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(),
        legend.position = "none")
```

### Investigating variable importance

Variables with treatment:
```{r var-imp, message = FALSE, warning = FALSE, fig.align = 'center'}
# occurrences of a variable in BART
treatment_interactions <-
  covariate_with_treatment_importance(te_model, treatment = "treatment")

treatment_interactions %>% 
  ggplot() + 
  geom_bar(aes(x = variable, y = avg_inclusion), stat = "identity") +
  theme_bw() + 
  ggtitle("Important variables interacting with treatment ('treatment')") + 
  ylab("Inclusion counts") +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

#### Variables overall:
```{r, message = FALSE, warning = FALSE, fig.align = 'center'}
variable_importance <-
  covariate_importance(te_model)

variable_importance %>% 
  ggplot() + 
  geom_bar(aes(x = variable, y = avg_inclusion), stat = "identity") +
  theme_bw() + ggtitle("Important variables overall") +
  ylab("Inclusion counts") +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```


## Regular Regression Tree

Following example from lecture 19 (<https://www.stat.cmu.edu/cmsac/sure/2022/materials/lectures/slides/19-Trees.html#1>)
```{r, message = FALSE, warning = FALSE, fig.align = 'center'}
library(rpart)
init_tree <- rpart(formula = ppg_21_22 ~ position + ppg_19_20 + treatment + gp_21_22 + age_continuous,
                   data = ohl_update, method = "anova")
```

Display tree
```{r, message = FALSE, warning = FALSE, fig.align = 'center'}
library(rpart.plot)
rpart.plot(init_tree)
plotcp(init_tree)
```
The bottom number tells you how much of the data is in a particular node. For example, 29% of players in ohl_update had a ppg in the 2019-2020 season less than 0.24 and played less than 60 games in the 2021-2022 season.


Full tree
```{r, message = FALSE, warning = FALSE, fig.align = 'center'}
full_tree <- rpart(formula = ppg_21_22 ~ position + ppg_19_20 + treatment + gp_21_22 + age_continuous,
                   data = ohl_update, method = "anova", control = list(cp = 0, xval = 10))
rpart.plot(full_tree)
plotcp(full_tree)
```
Not fully sure how to interpret this graph.

Train with caret:
```{r, message = FALSE, warning = FALSE, fig.align = 'center'}
# library(caret)
# caret_tree <- train(formula = ppg_21_22 ~ position + ppg_19_20 + treatment + gp_21_22 + age_continuous+ pts_19_20 + pm_19_20,
#                    data = ohl_update, method = "rpart", trControl = trainControl(method = "cv", number = 10), tuneLength = 20)
# ggplot(caret_tree) + theme_bw()
```

Error in na.fail.default(list(season = c("2019-2020", "2019-2020", "2019-2020", : 
missing values in object

